


\chapter{Complete Algorithm}
\label{sec:CompleteAlgorithm}

Once the fundamental situational evaluation functions are established, they can be combined to produce a complete betting algorithm.
The poker bot searches for its optimal bet by defining an overall scalar utility function, and performing typical 1-D function optimization to find the global maximum (e.g. golden section search with parabolic interpolation, etc.).
As long as the utility function is sufficiently close to unimodal, the optimal bet amount can always be located efficiently.
The following sections describe the mathematics and implementation of the overall betting algorithm.

\section{Overall Utility}



\subsection{Outcome Expectation}
\label{sec:StateModel}

The overall utility function weighs the utility of each possible outcome with the probability of achieving that outcome.
For example, if $\mathrm{Pr\{push}\}$ and $\mathrm{Pr\{raised}\}$ are known, %$\mathrm{E[callers}_x]$, 
\[
U_{\mathrm{StateModel}} = \left(U_{\mathrm{showdown}}\right)^{\mathrm{Pr\{call\}}} \cdot \left(U_{\mathrm{push}}\right)^{\mathrm{Pr\{push}\}} \cdot \prod_{\mathrm{raiseAmount}} \left(U_{\mathrm{raiseAmount}}\right)^{\mathrm{Pr\{raiseAmount}\}}
\]
Here, $\mathrm{Pr\{call\}}$ is simply $\left(1 - \mathrm{Pr\{push}\} - \sum \mathrm{Pr\{raised}\}\right)$.
Again, all utilities and probabilities are functions of $b$, the size you choose to bet.
Ultimately a 1-D function extremum search takes place across $U\left(b\right)$.

Note that because it is too computationally expensive to consider all possible $raiseAmount$ values, an approximation is employed.
As implemented, the $raiseAmount$ variable iterates across an array of select values that opponents could raise to.
The set of possible raises is a logarithmic quantization of the full range of raise amounts; $U_{\mathrm{raiseAmount}}$ and $\mathrm{Pr\{raised}\}$ are precomputed at each of the quantized values.
The \texttt{StateModel} class implements this final aggregation of outcomes.

\subsection{Geometric vs. Arithmetic Utility}
\label{sec:GeomAlgb}

For small bets, the ideal utility function is based on the geometric mean utility from section~\ref{sec:Utility}.
For large bets, the arithmetic mean equivalent is preferred.
The actual utility function is therefore a linear combination of both the geometric and arithmetic utility functions:
\[
U_{\mathrm{combined}}(b) = \left(1 - \frac{b}{riskprice}\right) U_{\mathrm{geom}}(b) + \left(\frac{b}{riskprice}\right) U_{\mathrm{algb}}(b)
\]
As $b$ increases toward $riskprice$, the weight shifts from $U_{\mathrm{geom}}(b)$ to $U_{\mathrm{algb}}(b)$.
Here $riskprice$ is the maximum feasible bet: either your total number of chips, or the smallest bet size for which the opponents' FoldGain become \emph{always} positive.
The \texttt{AutoScalingFunction} class provides this linear combination framework for combining utility functions.

\subsection{Worst Case Win Percentage}
\label{sec:statworse}

For small bets $\mathrm{Pr\{push\}}$ will tend to be low, and based on the histogram of possible opponents $\mathrm{Pr\{win\}}$ and $\mathrm{Pr\{best\}}$ will be relatively accurate.
%the probabilty of winning the showdown (Section~\ref{sec:Mean}) as well as the probability of having the best hand (Section~\ref{sec:Rank}) will be relatively accurate.
For large bets where $\mathrm{Pr\{push\}}$ is higher, most opponents will choose to fold, and only the $\left(1 - \mathrm{Pr\{push\}} \right)$ best opponents are likely to continue through to a showdown.
In the extreme case, $\mathrm{Pr\{push\}} \rightarrow 1$ and only the best few hands will even consider playing.
When this happens, $\mathrm{Pr\{win\}}$ should approach the worst case win percentage: the chance of winning against the strongest opposing hand.

Unfortunately, recomputing $\mathrm{Pr\{win\}}$ for every $\mathrm{Pr\{push\}}$ is computationally expensive.
Therefore, we apply the same linear combination strategy as used to balance the geometric vs. arithmetic utility.
Two separate utility functions are created: one with the standard $\mathrm{Pr\{win\}}$, and one with the worst case $\mathrm{Pr\{win\}}$ (denoted with the subscript `fear').
These two utility functions are then dynamically weighted in a linear combination based on the bet size $b$:
\[
U_{\mathrm{combined}}(b) = \left(1 - \frac{b - b_{\mathrm{min}}}{riskprice - b_{\mathrm{min}}}\right) U_{\mathrm{main}}(b) + \left(\frac{b - b_{\mathrm{min}}}{riskprice - b_{\mathrm{min}}}\right) U_{\mathrm{fear}}(b)
\]
Here, $b_{\mathrm{min}}$ is the largest bet where $Pr\{push\}$ is still zero.
If $b < b_{\mathrm{min}}$, only $U_{\mathrm{main}}(b)$ is used.

\section{Implementation}
\label{sec:CompleteImplementation}
To assist the reader in browsing through the source code for this project, the main entrypoints to the key algorithms are listed below.
The core poker bot betting algorithms are implemented along the following code path:
\begin{enumerate}
\singlespacing
\item \texttt{PositionalStrategy::StoreDealtHand(const CommunityPlus \& o)}

Calling this method assigns a new hand to a bot/player.
Bots will later call \texttt{ViewDealtHand()} to access their current hand.

\item \texttt{PositionalStrategy::SeeCommunity(const Hand\& h, const int8 cardsInCommunity)}

Calling this method informs the bot that the next betting round is about to begin.
It will now have all the information that it needs in order to begin histogram generation.
Histogram generation requires a significant portion of runtime; by generating the histograms ahead of time, less computation will required when the poker bot's turn to bet begins.
Furthermore, a bot may need to bet multiple times in a betting round, but since the community cards won't change in the middle of a betting round, there is no need to recompute any of the histograms until the next betting round is about to begin.

\item \texttt{ImproveGainStrategy::MakeBet()} and \texttt{DeterredGainStrategy::MakeBet()}

Calling the \texttt{MakeBet()} function initiates the poker bot's search for the optimal bet in the current situation.
The bot applies a unimodal function extremum search to identify the bet size that leads to the best overall utility.
Upon completion, these functions return the optimal bet size as determined by the poker bot.
\end{enumerate}

