
\clearpage


\begin{flushleft}

\chapter*{Glossary}
 \addcontentsline{toc}{chapter}{Glossary}

\paragraph*{Betting Round} A betting round is a sequence of bets that takes place until all players have folded or called.

\paragraph*{Call} A bet that is the same amount as the current betting round's highest bet so far.
 
\paragraph*{Community Cards} The community cards are the cards dealt directly to the table. These cards are equally available to all players that haven't folded, and make up five of the seven cards a player can choose from during the showdown.
 
\paragraph*{Five-card Poker Hand} During the showdown, players attempt to win the pot by forming the strongest possible five card poker hand out of the seven card hand that they have available.
 
\paragraph*{Flop} The first set of community cards dealt to the table. The flop is a set of three cards.

\paragraph*{Fold} Forfieting your hole cards and your chance to win any chips for the remainder of the game round.

\paragraph*{Game Round} A game round is a sequence of betting rounds, possibly culminating in a showdown.
 
\paragraph*{Hole Cards} Hole cards are the two cards a player is dealt to begin each game round.
 
\paragraph*{Post-flop} The second betting round, after the flop is dealt to the table.

\paragraph*{Pot} The pot is the entire sum of chips that can be won by a player during a game round. Betting money contributes to the pot, and the total pot is composed of the sum of all bets made throughout the current game round by all players.
%total collection of chips that have been bet so far throughout the entire game round. 
 
\paragraph*{Pre-flop} The first betting round, before the flop is dealt to the table.
 
\paragraph*{Raise} A bet that is a greater amount than the current betting round's highest bet so far.

\paragraph*{Re-raise} A raise that is made in response to someone else's raise during the same betting round.
 
\paragraph*{Ring Game} In a poker tournament, it costs a single flat fee to enter, and the prize money is distributed based on the rank that the players finish in. In a ring game, players bring their own money to the table, any money won or lost at the table is kept, usually with no predetermined time limit.
 
\paragraph*{River} The third set of community cards dealt to the table. The river is one card.

\paragraph*{Seven-card hand} When the showdown is reached, each player will have two hole cards and the table will show five community cards.

\paragraph*{Showdown} After all the betting rounds have completed, if there is still more than one player who hasn't folded, cards are turned face up to decide the winner. The winner of a showdown is the player who can form the strongest five card hand.

\paragraph*{Split Pot} If, after all tiebreakers, two players still have the same five-card hand strength, the winnings are split between both players. A split pot can also occur with three or more players.

\paragraph*{Turn} The second set of community cards dealt to the table. The turn is one card.





\end{flushleft}


%%%%%%%%%%%%%%% Bibliography

\begin{flushleft}
\bibliographystyle{ieeetr}
\bibliography{document/references}

\addcontentsline{toc}{chapter}{References}
\end{flushleft}
\clearpage

